---
title: "R PostgreSQL Notes"
output: rmdformats::material
---
# Steps to Setup PostgreSQL

I decided to load spatial data for my Portland Tree map from a database rather than relying on individual files. These are the steps I took to set up PostgreSQL and PostGIS on the same server as Shiny. There were a couple of roadblocks along the way that I want to document solutions to.

*Note: The app uses fairly stable data and was previously loaded from a geopackage (.gpkg) which is a database. While a PostgreSQL database is unneeded for data of this size I wanted to go through the setup steps*

First load a few libraries for interacting with the database and loading spatial data.

```{r setup, message=FALSE}
# For reading spatial data
library(sf)
# Database connections
library(odbc)
library(DBI)
# Postgre specific database connections
library(RPostgreSQL)

knitr::opts_chunk$set(eval = F)
```

1) Install PostgreSQL on the digital ocean server.
This is fairly straightforward but since I'm using DO they have [nice help documents](https://www.digitalocean.com/community/tutorials/how-to-install-and-use-postgresql-on-ubuntu-20-04).
`sudo apt install postgresql postgresql-contrib`

Once it was installed I created a new user and database.
  - Use `sudo -u postgres createuser --interactive`
  - Create user `max` and database `parks`
  - Connect to `parks` database from DO with `psql -d parks`

2) Change the default password

Always a good idea to set a new password for something that is on the internet. I followed [this link](https://chartio.com/resources/tutorials/how-to-set-the-default-user-password-in-postgresql/#changing-the-password)

3) Install PostGIS

- I used [this link for reference](https://computingforgeeks.com/how-to-install-postgis-on-ubuntu-debian/), but the steps are essentially to log into your postgres database that you just made and enable postgis once its installed.

```{bash, eval = F}
# Update to your version of postgres
sudo apt install postgis postgresql-{{postgres-version}}-postgis-{{postgis-version}}

# log into your postgres database
psql -d {yourdb}
CREATE EXTENSION postgis;
```

Things are mostly set up at this point. The following points were things I changed to be able to connect to the database and get things running.

4) Set Firewall to Allow Connections

PostgreSQL dwfault port is 5432 and you will need to open this up to be able to connect to the database. I had `UFW` previously installed for firewall purposes so I used this to change the permissions. 

First check the firewall status
```{bash, eval = F}
sudo ufw status
```

If not enabled
```{bash, eval = F}
sudo ufw enable
```

Then add port 5432 and reload `UFW` take enable the new rule.
```{bash, eval = F}
sudo ufw allow 5432/tcp
sudo ufw reload
```

[This guide](https://www.vultr.com/docs/how-to-configure-ufw-firewall-on-ubuntu-14-04) goes over the steps needed plus more.

**Try connecting to the database at this point from a local machine and through Shiny**

5) Change Config Files

This tripped me up for some time. I was still getting connection issues from Shiny even though I was able to connect locally. If you are able to connect then great! You're all done. If not there is probably an issue with permission protocols. I was specifically getting the error 'Peer authentication failed for user “postgres” (or any user)' which led me [this solution](https://gist.github.com/AtulKsol/4470d377b448e56468baef85af7fd614). 

Essentially navigate to your `pg_hba.conf` file, *mine was at: `/etc/postgresql/13/main/pg_hba.conf`*, and change 
```{bash, eval = F}
# TYPE DATABASE USER ADDRESS METHOD
local  all      all          peer
```

to 
```{bash, eval = F}
# TYPE DATABASE USER ADDRESS METHOD
local  all      all          md5
```

**Pay attention to the details of the above lines.** I changed a similar but different line in the config file,
```{bash, eval = F}
local   all   postgres        md5
```

which led this solution to not work and caused an unneeded multi-hour search into trying to solve the original error. Finally restart postgresql with `sudo service postgresql restart`

# Load Data into Database

Now that everything is finally set up we can load data into the remote database. I previously had a few spatial layers stored in a geopackage (*which actually works pretty well for this type of application*).

```{r, eval=F}
# get all layers
layers <- st_layers("../../tree_map/dat/data.gpkg")

# read all layers
lays <- sapply(layers$name, function(x) read_sf("../../tree_map/dat/data.gpkg", layer = x))
```

Create the connection to the database we just set up.

```{r, eval=F}
con <- dbConnect(PostgreSQL(),
     , host = Sys.getenv("postgre_ip")
     , port = '5432'
     , dbname = 'parks'
     , user = 'max'
     , password = Sys.getenv("postgre_pswd"))

# list the current tables
# This should return just a default spatial reference layer created by Postgis
dbListTables(con)
```

### A note about .Renviron
You'll notice the use of `Sys.getenv()` for the host name and password. This has two uses in the database connection.

1) It's a bad idea to store a password in open code that anyone can read. I have a `.Renviron` file, that is not tracked with git, in my Rproject folder that contains variables 'postgre_ip' and 'postgre_pswd'. The `Sys.getenv()` calls these variables so they never need to written out. This is also useful for API keys.

2) Since I am developing locally and running the app from a VM I can use this method have different login information for my server. On the server I have a different `.Renviron`,again not tracked by git, which has a different IP address. I could also vary the user and password if there were permissions I wanted to track.

**Note:** The `.Renviron` file that gets loaded by shiny-server is in `/home/shiny/` as opposed to the Rproject folder. There may be a way to change this.

If you use the `.Renviron` method make sure to restart shiny server since it reads that file on boot.
```{bash}
sudo restart shiny-server
```


Once the connection is set up you can write files to the database directly with the `sf` package

```{r}
# Write each file
st_write(obj = lays[["park"]], dsn = con, drop = T, layer = "park")
st_write(obj = lays[["neigh"]], dsn = con, layer = "neigh")
st_write(obj = lays[["portland"]], dsn = con, layer = "portland")
st_write(obj = lays[["park_trees"]], dsn = con, layer = "park_trees")

# check tables
dbListTables(con)
```

## Test Connection

Read a file just to make sure everything works

```{r}
read_sf(con, "portland")
```

